A Hybrid Neutronics Method for Time-Dependent Control Rod Modeling in Molten Salt Reactors

### 2024-06-15 hybrid.tex
hybrid method edits

Section \ref{sec:summary-nts-mtds} highlighted the poor performance of neutron diffusion
methods for calculating neutron fluxes near control rods. Strong neutron absorption in the control
rod region produces a highly anisotropic neutron flux extending some distance outside the control
rod. Neutron transport methods, which retain angular dependence of the neutron flux to various
extents, generally fare better than neutron diffusion methods, which have isotropic diffusion
coefficients. However, neutron transport methods are also generally more computationally expensive,
given the increased dimensionality of the problem from the angular component. Adding an angular
dimension to the existing geometric and neutron energy group dimensions dramatically
increases the problem size and the computational resources necessary to solve the system. Many past
efforts have tried introducing
transport correction techniques to improve neutron flux and multiplication factor estimates in
diffusion-based methods. Other than control rod regions, these techniques may also correct
homogenization errors introduced by spatial homogenization of fuel assemblies and other
structures within a reactor core. They invariably rely on neutron transport methods to generate
transport corrections in the form of corrected diffusion coefficients
\cite{bretscher_computing_1997, scherer_determination_1976, ronen_accurate_2004,
pounders_diffusion_2009, kavenoky_sph_1978}, boundary conditions \cite{davison_influence_1951,
pellaud_extrapolation_1968, fen_modelling_1992}, Eddington factors, or discontinuity factors
\cite{koebke_new_1980}.

In this chapter, I propose a hybrid neutronics method for improving control rod modeling in neutron
diffusion solvers without spatial homogenization. In essence, the hybrid method involves applying
the $S_N$ discrete ordinates neutron transport method on subregions containing the control rod to
obtain pointwise transport corrections for the diffusion method on the subregions.

Section \ref{sec:theory} discusses the theoretical background for the hybrid $S_N$-diffusion
method. Section \ref{sec:implementation} provides numerical implementation details of
the hybrid method and its components. Sections \ref{sec:test-case} and \ref{sec:sim-param} describe
the 1-D test cases and several simulation parameters in that context. Section
\ref{sec:prelim-results} discusses the results of the hybrid method applied to the 1-D test cases
with comparisons to higher-fidelity Monte Carlo and $S_N$ neutron transport methods. Lastly,
Section \ref{sec:hybrid-summary} summarizes the key findings in this chapter.

\section{Theory} \label{sec:theory}

The proposed hybrid $S_N$-diffusion method is an iterative method to improve the
accuracy of neutron diffusion solutions in reactor systems with highly anisotropic flux regions
which I will use to model control rods. This section presents the mathematical derivation for the
\gls{SAAF} formulation of the $S_N$ equations, the neutron diffusion equations, the drift
transport correction term, and the computational algorithm for the hybrid method.

\subsection{\gls{SAAF} Formulation of the $S_N$ Method}

\subsubsection{Multigroup Neutron Transport Equations}

@@

  \bm{\Psi} =

@@

  \bm{\phi} = \int_S \bm{\Psi}d\hat{\Omega} =

@@

  \bm{\frac{\Psi}{v}}

@@

  \bm{Q}^{\text{ext}} =

@@

  \bm{Q}_f = \frac{1}{4\pi}\bm{Q}_{f,0} =
    \frac{1}{4\pi}\chi_{p,1}(1-\beta)\sum^G_{g'=1} \nu\Sigma_{f,g'} \phi_{g'} +
    \frac{1}{4\pi}\chi_{p,2}(1-\beta)\sum^G_{g'=1} \nu\Sigma_{f,g'} \phi_{g'} +
    \frac{1}{4\pi}\chi_{p,G}(1-\beta)\sum^G_{g'=1} \nu\Sigma_{f,g'} \phi_{g'} +
  \bm{Q} = \bm{Q}_f + \bm{Q}^\text{ext}. \nonumber

@@

  \mathbb{L}_1\bm{\Psi} =

@@

  \mathbb{L}_2\bm{\Psi} =

@@

  \mathbb{L}\bm{\Psi} = \mathbb{L}_1\bm{\Psi} + \mathbb{L}_2\bm{\Psi}, \nonumber \\
  \mathbb{S}\bm{\Psi} =

@@

%  \mathbb{F}_p\bm{\Psi} = \frac{1}{4\pi} \mathbb{F}_{p,0}\bm{\phi} =
%    \frac{1}{4\pi}\chi_{p,1}(1-\beta)\sum^G_{g'=1}\nu\Sigma_{f,g'}\phi_{g'} \\
%    \frac{1}{4\pi}\chi_{p,2}(1-\beta)\sum^G_{g'=1}\nu\Sigma_{f,g'}\phi_{g'} \\
%    \frac{1}{4\pi}\chi_{p,G}(1-\beta)\sum^G_{g'=1}\nu\Sigma_{f,g'}\phi_{g'}
  \mathbb{B}\bm{\Psi} =

@@

@@

  = \mathbb{S}\bm{\Psi} + \bm{Q}_f + \bm{Q}^\text{ext} \label{eq:nte-vec}

@@

  \bm{\Psi} = \bm{\Psi}^\text{inc} + \mathbb{B}\bm{\Psi}

@@

\subsubsection{Weak Form of the Multigroup Neutron Transport Equations}

@@

\subsubsection{\gls{SAAF} Formulation}

@@

\subsubsection{Spatial Discretization and Void Treatment}

@@

\subsubsection{Angular Discretization}

The $S_N$ method applies a collocation method for the angular discretization by solving for the
angular flux along $N_d$ discrete directions $\hat{\Omega}_d$ with weights $w_d$. Numerous options
exist for the choice of the angular quadrature set
$\{\hat{\Omega}_d,w_d | d=1,\dots,N_d\}$.
For this work, we chose the commonly used level-symmetric quadrature set
\cite{wang_rattlesnake_2018} for ease of implementation, particularly when applying reflective
boundary conditions due to 90-degree rotational symmetry about each of the three Cartesian axes.

We use the level-symmetric quadrature sets to approximate integrals of the angular flux over the
angular domain
%
\begin{gather}
  \phi_g = \int_\mathcal{S} \Psi_{g,d}\hat{\Omega} \approx \sum^{N_d}_{d=1}w_d
  \Psi_{g,d}), \label{eq:0th-mmt}
  \shortintertext{where}
  \Psi_{g,d} = \Psi_g(\hat{\Omega}_d). \nonumber
\end{gather}
%
We calculate the scalar flux $\phi_g$ by taking the zeroth moment with respect to $\hat{\Omega}$ as
shown in Eq.\ \ref{eq:0th-mmt}. We can also calculate higher moments
%
\begin{gather}
  \phi_{g,l,m} = \int_\mathcal{S} Y_{l,m}(\hat{\Omega})\Psi_g(\hat{\Omega})d\hat{\Omega} \approx
  \sum^{N_d}_{d=1}w_d Y_{l,m}(\hat{\Omega}_d)\Psi_{g}(\hat{\Omega}_d) \label{eq:nth-mmt}
  \shortintertext{where}
  \begin{align*}
    Y_{l,m}(\mu,\omega) &=
    \begin{cases}
      \sqrt{2}C^{|m|}_l P^{|m|}_l \cos(|m|\omega), & \mbox{ for } 0<m\leq l \\
      C^0_l P^0_l(\mu), & \mbox{ for } m=0, 0\leq l < \infty \\
      \sqrt{2} C^{|m|}_l P^{|m|}_l(\mu) \sin(|m|\omega), & -l \leq m < 0
    \end{cases}, \\
    \mu &= \mbox{cosine of the polar angle,} \\
    \omega &= \mbox{the azimuthal angle,} \\
    C^{|m|}_l &= \sqrt{\frac{\left(l-|m|\right)!}{\left(l+|m|\right)!}}, \\
    P^m_l &= \mbox{associated Legendre polynomial of degree $l$ and order $m$.}
  \end{align*}
\end{gather}
%
for anisotropic scattering treatment.

\subsubsection{Final Forms of the \gls{SAAF} $S_N$ Equations}

@@

Time derivative term:
%
\begin{gather}
  \left(\left(\mathbb{I}+\bm{\tau}\mathbb{L}_1\right)\bm{\Psi}^*,
  \frac{\partial}{\partial t}\left(\frac{\bm{\Psi}}{\bm{v}}\right)\right) =
  \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d}+\tau_g\hat{\Omega}_d\cdot\Psi^*_{g,d},
  \frac{\Psi_{g,d}}{v_g}\right)_\mathcal{D}
\end{gather}
%
Streaming term:
%
\begin{gather}
  \left(\mathbb{L}_1\bm{\Psi}^*,
  \left(\bm{\tau}\mathbb{L}_1-\mathbb{I}+\bm{\tau}\mathbb{L}_2\right)\bm{\Psi}\right) =
  \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\hat{\Omega}_d\cdot\Psi^*_{g,d},\tau_g\hat{\Omega}\cdot
  \nabla\Psi_{g,d}-(1-\tau_g\Sigma_{t,g})\Psi_{g,d}\right)_\mathcal{D}
\end{gather}
%
Collision term:
%
\begin{gather}
  \left(\mathbb{L}_2\bm{\Psi}^*,\bm{\Psi}\right) =
  \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d},\Sigma_{t,g}\Psi_{g,d}\right)_\mathcal{D}
\end{gather}
%
Scattering term:
%
\begin{gather}
  \left(\left(\mathbb{I}+\bm{\tau}\mathbb{L}_1\right)\bm{\Psi}^*,\mathbb{S}\bm{\Psi}\right) =
  \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d}+\tau_g\hat{\Omega}_d\cdot\Psi^*_{g,d},
    \sum^G_{g'=1}\sum^L_{l=0}\Sigma^{g'\rightarrow g}_{s,l}\sum^l_{m=-l}

>>>>

Section \ref{sec:summary-nts-mtds} highlights the poor performance of neutron diffusion
methods for calculating neutron fluxes near control rods. Strong neutron absorption in the control
rod region produces a highly anisotropic neutron flux extending some distance outside the control
rod. Neutron transport methods, which retain angular dependence of the neutron flux to various
extents, generally fare better than neutron diffusion methods, which have isotropic diffusion
coefficients. However, neutron transport methods are also generally more computationally expensive,
given the increased dimensionality of the problem from the angular component. Adding an angular
dimension to the existing geometric and neutron energy group dimensions dramatically
increases the problem size and the computational resources necessary to solve the system. Many past
efforts have tried introducing
transport correction techniques to improve neutron flux and multiplication factor estimates in
diffusion-based methods. Other than control rod regions, these techniques may also correct
homogenization errors introduced by spatial homogenization of fuel assemblies and other
structures within a reactor core. They invariably rely on neutron transport methods to generate
transport corrections in the form of corrected diffusion coefficients
\cite{bretscher_computing_1997, scherer_determination_1976, ronen_accurate_2004,
pounders_diffusion_2009, kavenoky_sph_1978}, boundary conditions \cite{davison_influence_1951,
pellaud_extrapolation_1968, fen_modelling_1992}, Eddington factors, or discontinuity factors
\cite{koebke_new_1980}.

In this chapter, I propose a hybrid $S_N$-diffusion neutronics method to improve control rod
modeling in neutron diffusion solvers without spatial homogenization. In essence, the hybrid
method is an iterative method that applies
the $S_N$ discrete ordinates neutron transport method on subregions containing the control rod to
obtain pointwise transport corrections for the diffusion method on the subregions.
This chapter presents the mathematical derivation for the
\gls{SAAF} formulation of the $S_N$ equations and the drift transport correction term for the
neutron diffusion equations, and the computational algorithm for the hybrid method.

% Section \ref{sec:theory} discusses the theoretical background for the hybrid $S_N$-diffusion
% method. Section \ref{sec:implementation} provides numerical implementation details of
% the hybrid method and its components. Sections \ref{sec:test-case} and \ref{sec:sim-param} describe
% the 1-D test cases and several simulation parameters in that context. Section
% \ref{sec:prelim-results} discusses the results of the hybrid method applied to the 1-D test cases
% with comparisons to higher-fidelity Monte Carlo and $S_N$ neutron transport methods. Lastly,
% Section \ref{sec:hybrid-summary} summarizes the key findings in this chapter.

% The proposed hybrid $S_N$-diffusion method is an iterative method to improve the
% accuracy of neutron diffusion solutions in reactor systems with highly anisotropic flux regions
% which I will use to model control rods.

\section{\gls{SAAF} Formulation of the $S_N$ Method}

\subsection{Multigroup Neutron Transport Equations}

@@

  \bm{\Psi} \equiv

@@

  \bm{\phi} \equiv \int_S \bm{\Psi}d\hat{\Omega} \equiv

@@

  \bm{\frac{\Psi}{v}} \equiv

@@

  \bm{C} \equiv
    C_1 \\
    C_2 \\
    C_G
  \bm{Q}^{\text{ext}} \equiv

@@

  \end{bmatrix}, \nonumber
%  \bm{Q}_f = \frac{1}{4\pi}\bm{Q}_{f,0} =
%    \frac{1}{4\pi}\chi_{p,1}(1-\beta)\sum^G_{g'=1} \nu\Sigma_{f,g'} \phi_{g'} +
%    \frac{1}{4\pi}\sum^I_{i=1}\chi_{d,1} \lambda_i C_i \\
%    \frac{1}{4\pi}\chi_{p,2}(1-\beta)\sum^G_{g'=1} \nu\Sigma_{f,g'} \phi_{g'} +
%    \frac{1}{4\pi}\sum^I_{i=1}\chi_{d,2} \lambda_i C_i \\
%    \frac{1}{4\pi}\chi_{p,G}(1-\beta)\sum^G_{g'=1} \nu\Sigma_{f,g'} \phi_{g'} +
%    \frac{1}{4\pi}\sum^I_{i=1}\chi_{d,G} \lambda_i C_i
%  \bm{Q} = \bm{Q}_f + \bm{Q}^\text{ext}. \nonumber

@@

  \mathbb{L}_1\bm{\Psi} \equiv

@@

  \mathbb{L}_2\bm{\Psi} \equiv

@@

  \mathbb{L}\bm{\Psi} \equiv \mathbb{L}_1\bm{\Psi} + \mathbb{L}_2\bm{\Psi}, \nonumber \\
  \mathbb{S}\bm{\Psi} \equiv

@@

  \mathbb{B}\bm{\Psi} \equiv

@@

  \mathbb{F}\bm{\Psi} \equiv \frac{1}{4\pi}\mathbb{F}_0\bm{\Psi} \equiv
    \frac{1}{4\pi}\chi_{p,1}(1-\beta)\sum^G_{g'=1}\nu\Sigma_{f,g'}\phi_{g'} \\
    \frac{1}{4\pi}\chi_{p,2}(1-\beta)\sum^G_{g'=1}\nu\Sigma_{f,g'}\phi_{g'} \\
    \frac{1}{4\pi}\chi_{p,G}(1-\beta)\sum^G_{g'=1}\nu\Sigma_{f,g'}\phi_{g'}
  \mathbb{C}\bm{C} \equiv \frac{1}{4\pi}\mathbb{C}_0\bm{C} \equiv

@@

  = \mathbb{S}\bm{\Psi} + \bm{Q}, \label{eq:nte-vec}

@@

  \bm{\Psi} = \bm{\Psi}^\text{inc} + \mathbb{B}\bm{\Psi},
    \bm{Q} &= \bm{Q}_f + \bm{Q}^\text{ext}, & \\
    \bm{Q}_f &= \mathbb{F}\bm{\Psi} + \mathbb{C}\bm{C}. &

@@

\subsection{Weak Form of the Multigroup Neutron Transport Equations}

@@

\subsection{\gls{SAAF} Formulation}

@@

\subsection{Spatial Discretization and Void Treatment}

@@

\subsection{Angular Discretization}

The $S_N$ method applies a collocation method for the angular discretization by solving for the
angular flux along $N_d$ discrete directions $\hat{\Omega}_d$ with weights $w_d$. Numerous options
exist for the choice of the angular quadrature set
$\{\hat{\Omega}_d,w_d\ |\ d=1,\dots,N_d\}$.
For this work, we chose the commonly used level-symmetric quadrature set
\cite{wang_rattlesnake_2018} for ease of implementation, particularly when applying reflective
boundary conditions due to 90-degree rotational symmetry about each of the three Cartesian axes.

We use the level-symmetric quadrature sets to approximate integrals of the angular flux over the
angular domain
%
\begin{gather}
  \phi_g = \int_\mathcal{S} \Psi_{g,d}\hat{\Omega} \approx \sum^{N_d}_{d=1}w_d
  \Psi_{g,d}, \label{eq:0th-mmt}
  \shortintertext{where}
  \Psi_{g,d} = \Psi_g(\hat{\Omega}_d), \nonumber \\
  w = \sum^{N_d}_{d=1} w_d\ (= 8 \mbox{ for 3-D level-symmetric set}). \nonumber
\end{gather}
%
We calculate the scalar flux $\phi_g$ by taking the zeroth moment with respect to $\hat{\Omega}$ as
shown in Eq.\ \ref{eq:0th-mmt}. We can also calculate higher moments
%
\begin{gather}
  \phi_{g,l,m} = \int_\mathcal{S} Y_{l,m}(\hat{\Omega})\Psi_g(\hat{\Omega})d\hat{\Omega} \approx
  \sum^{N_d}_{d=1}w_d Y_{l,m}(\hat{\Omega}_d)\Psi_{g}(\hat{\Omega}_d) \label{eq:nth-mmt}
  \shortintertext{where}
  \begin{align*}
    Y_{l,m}(\mu,\omega) &=
    \begin{cases}
      \sqrt{2}C^{|m|}_l P^{|m|}_l \cos(|m|\omega), & \mbox{ for } 0<m\leq l \\
      C^0_l P^0_l(\mu), & \mbox{ for } m=0, 0\leq l < \infty \\
      \sqrt{2} C^{|m|}_l P^{|m|}_l(\mu) \sin(|m|\omega), & -l \leq m < 0
    \end{cases}, \\
    \mu &= \mbox{cosine of polar angle,} \\
    \omega &= \mbox{azimuthal angle,} \\
    C^{|m|}_l &= \sqrt{\frac{\left(l-|m|\right)!}{\left(l+|m|\right)!}}, \\
    P^m_l &= \mbox{associated Legendre polynomial of degree $l$ and order $m$.}
  \end{align*}
\end{gather}
%
for anisotropic scattering treatment.

\subsection{Final Forms of the \gls{SAAF} $S_N$ Equations}

@@

\noindent Time derivative term:
%
\begin{gather}
  \left(\left(\mathbb{I}+\bm{\tau}\mathbb{L}_1\right)\bm{\Psi}^*,
  \frac{\partial}{\partial t}\left(\frac{\bm{\Psi}}{\bm{v}}\right)\right) =
  \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d}+\tau_g\hat{\Omega}_d\cdot\Psi^*_{g,d},
  \frac{\Psi_{g,d}}{v_g}\right)_\mathcal{D}
\end{gather}
%
Streaming term:
%
\begin{gather}
  \left(\mathbb{L}_1\bm{\Psi}^*,
  \left(\bm{\tau}\mathbb{L}_1-\mathbb{I}+\bm{\tau}\mathbb{L}_2\right)\bm{\Psi}\right) =
  \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\hat{\Omega}_d\cdot\nabla\Psi^*_{g,d},\tau_g\hat{\Omega}
  \cdot\nabla\Psi_{g,d}-(1-\tau_g\Sigma_{t,g})\Psi_{g,d}\right)_\mathcal{D}
\end{gather}
%
Collision term:
%
\begin{gather}
  \left(\mathbb{L}_2\bm{\Psi}^*,\bm{\Psi}\right) =
  \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d},\Sigma_{t,g}\Psi_{g,d}\right)_\mathcal{D}
\end{gather}
%
Scattering term:
%
\begin{gather}
  \left(\left(\mathbb{I}+\bm{\tau}\mathbb{L}_1\right)\bm{\Psi}^*,\mathbb{S}\bm{\Psi}\right) =
  \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d}+\tau_g\hat{\Omega}_d\cdot\nabla\Psi^*_{g,d},
  \sum^G_{g'=1}\sum^L_{l=0}\Sigma^{g'\rightarrow g}_{s,l}\sum^l_{m=-l}
  \frac{2l+1}{w}Y_{l,m}(\hat{\Omega}_d)\phi_{g',l,m}\right)_\mathcal{D}
\end{gather}
%
Prompt fission source term:
%
\begin{gather}
  \left(\left(\mathbb{I}+\bm{\tau}\mathbb{L}_1\right)\bm{\Psi}^*,\mathbb{F}\bm{\Psi}\right) =
  \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d}+\tau_g\hat{\Omega}_d\cdot\nabla\Psi^*_{g,d},
  \frac{1}{w}\bar{\chi}_g\sum^G_{g'=1}\nu\Sigma_{f,g'}\phi_{g'}\right)_\mathcal{D}
\end{gather}
%
where $\bar{\chi}_g=\chi_{p,g}\left(1-\beta\right)$ for time-dependent problems with \glspl{DNP}
and $\bar{\chi}_g=\chi_{g}$ for steady-state problems without \glspl{DNP}.

\noindent Delayed neutron source term:
%
\begin{gather}
  \left(\left(\mathbb{I}+\bm{\tau}\mathbb{L}_1\right)\bm{\Psi}^*,\mathbb{C}\bm{C}\right) =
  \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d}+\tau_g\hat{\Omega}_d\cdot\nabla\Psi^*_{g,d},
  \frac{1}{w}\sum ^I_{i=1}\chi_{d,g}\lambda_i C_i\right)_\mathcal{D}
\end{gather}
%
We scale both prompt fission and delayed neutron source terms by the inverse of the
multiplication factor $\frac{1}{k}$ for $k$-eigenvalue problems.

\noindent Vacuum boundary term:
%
\begin{gather}
  \langle\bm{\Psi}^*,\bm{\Psi}\rangle^+ - \langle\bm{\Psi}^*,\bm{\Psi}^\text{inc}\rangle^- =
  \begin{cases}
    \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d},
    \hat{\Omega}_d\cdot\hat{n}_b\Psi_{g,d}\right),
    & \hat{\Omega}\cdot\hat{n}_b>0,\vec{r}\in\partial\mathcal{D}_s \\
    \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d},
    \hat{\Omega}_d\cdot\hat{n}_b\Psi^\text{inc}_{g,d}\right),
    & \hat{\Omega}\cdot\hat{n}_b<0,\vec{r}\in\partial\mathcal{D}_s
  \end{cases},
\end{gather}
%
where $\Psi^\text{inc}_{g,d}=0$ for a vacuum boundary and positive for a boundary source.

\noindent Reflecting boundary term:
%
\begin{gather}
  \langle\bm{\Psi}^*,\bm{\Psi}\rangle^+ - \langle\bm{\Psi}^*,\mathbb{B}\bm{\Psi}\rangle^- =
  \begin{cases}
    \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d},
    \hat{\Omega}_d\cdot\hat{n}_b\Psi_{g,d}\right),
    & \hat{\Omega}\cdot\hat{n}_b>0,\vec{r}\in\partial\mathcal{D}_s \\
    \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d},
    \hat{\Omega}_d\cdot\hat{n}_b\Psi_{g,d_r}\right),
    & \hat{\Omega}\cdot\hat{n}_b<0,\vec{r}\in\partial\mathcal{D}_s
  \end{cases},
  \shortintertext{where}
  \hat{\Omega}_{d_r} = \hat{\Omega}_d - 2(\hat{\Omega}_d\cdot\hat{n}_b)\hat{n}_b. \nonumber
\end{gather}

This concludes the derivation of the \gls{SAAF} formulation of the $S_N$ neutron transport method
for generating transport corrections in the hybrid $S_N$-diffusion method.

\section{Transport Correction Formulation}

The conventional approach for determining diffusion coefficients for each subregion involves
running a high-fidelity neutron transport simulation to tally region-wide estimates of the neutron
transport cross section. The transport cross section formulation is derived from the $P_1$
approximation of the neutron transport equation with isotropic sources \cite{bell_nuclear_1970} as
shown in Eq.\ \ref{eq:p1-diffcoef}.
Essentially, each defined subregion has a constant diffusion coefficient value. However, as
discussed in Section \ref{sec:summary-nts-mtds}, the neutron diffusion
equation is only valid in regions of high scattering-to-removal ratios with, at most, linearly
anisotropic scattering and small flux gradients. These conditions do not hold near or within
control rods, near interfaces of neighboring materials with highly dissimilar neutronic properties,
and in materials with significant scattering contributions from light nuclei.

Three main options exist in the literature for applying
transport corrections to the neutron diffusion equations: applying
diffusion coefficient corrections, adding a drift term, or adding a
correction term.

\subsection{Diffusion Coefficient Correction}

Diffusion coefficient corrections involve replacing the diffusion coefficient $D_g$ in the
diffusion term with ``optimal'' diffusion coefficients based on
pointwise transport corrections. The literature review in Chapter \ref{chap:lit} covers two such
examples in the Ronen method by Ronen \cite{ronen_accurate_2004} and the space-dependent diffusion
coefficients by Pounders \& Rahnema \cite{pounders_diffusion_2009}.

For the hybrid $S_N$-diffusion method, I used a formulation that incorporates pointwise
corrections to the neutron diffusion flux solution from the $S_N$-derived flux solution as follows:
%
\begin{align}
  D^s_g(x) &= -J^{tr}_g(x)\bigg/\frac{d\phi^{tr}_g(x)}{dx}. \label{eq:svdc}
\end{align}
%
where $D^s_g$ is the \gls{SVDC}, and the $tr$ superscript denotes the transport-derived neutron
current and scalar flux solutions from the $S_N$ method.
The basic form of this equation is Fick's first law of diffusion. The result is a diffusion
coefficient variable that varies in space from point to point. We assume that it is continuous and
at least once differentiable except at dissimilar material interfaces. \glspl{SVDC} provide
pointwise corrections to closely match the diffusion flux
solution to the $S_N$ flux solution.

Eq.\ \ref{eq:svdc} is identical to Ronen's \cite{ronen_accurate_2004} and Pounders \& Rahnema's
\cite{pounders_diffusion_2009} formulations for space-dependent diffusion coefficients in Eq.\
\ref{eq:ronen} and Eq.\ \ref{eq:emp}, respectively. In comparison with
the Ronen method, our approaches differ in how $\phi_{tr}$ and $J_{tr}$ are obtained. Starting with
a standard neutron diffusion calculation, the Ronen method applies an analytically derived
transport operator on every iteration to calculate a new estimate of $J_{tr}$ using the $\phi$
solution from the previous iteration. The main difficulty for the Ronen method lies in deriving the
transport operators which has been demonstrated for only 1-D geometries thus far.

For the empirical method developed by Pounders \& Rahnema, they assumed a priori knowledge of the
reference flux solution and used it to generate piecewise-constant, space-dependent diffusion
coefficients.
Pounders \& Rahnema \cite{pounders_diffusion_2009} demonstrated the effectiveness of applying
pointwise corrections derived from analytical or Monte Carlo reference flux solutions. Compared
with conventional $P_1$-based out-scatter and flux-limited approximations of the diffusion
coefficient, their diffusion coefficients showed superior agreement
with the reference flux solutions. They recognized that volume averaging within each mesh element
for piecewise-constant diffusion coefficients introduces some truncation error if the flux is
non-linear within the mesh element. This design choice may be due to an intention to retain the
diffusion coefficient as a constant in the $\frac{d}{dx}D\frac{d\phi}{dx}$ term of the neutron
diffusion equation (Eq.\ \ref{eq:1d-diff}).

Unlike their approach, the \gls{SVDC} formulation in Eq.\ \ref{eq:svdc} allows for continuously
varying diffusion coefficients to reduce truncation error. In practice, the discretization order of
\gls{SVDC} variables in a numerical calculation would be the same as the discretization order of
the reference flux solution. This formulation introduces a minor change to the finite difference
implementation of the 2nd-order diffusion term to handle spatial derivatives of the
diffusion coefficient as $D^s_g$ is not uniform in space.

Regardless of the differences in implementations amongst the methods discussed here,
diffusion coefficient corrections applied through Eq.\ \ref{eq:saaf} were
found to be effective in enabling the neutron diffusion method to accurately reproduce reference
flux solutions \cite{tomatis_application_2011, gross_high-accuracy_2020, tomatis_ronen_2021,
gross_comprehensive_2023, cufe_study_2024, pounders_diffusion_2009}.
However, a significant challenge for space-dependent diffusion coefficients
involves their resolution near neutron flux peaks. The neutron current and flux gradient values in
Eq.\ \ref{eq:svdc} generally do not reach zero at the same points in space,
resulting in diffusion coefficient values tending to positive or negative infinity when the
flux gradient is close to zero. Pounders \& Rahnema avoided this issue by using
larger mesh sizes to calculate their empirical diffusion coefficients. However, their
remedy contradicts mesh convergence requirements and would worsen flux accuracy in regions with
steep, non-linear fluxes, such as near control rods. For the Ronen method, Gross et al.\
\cite{gross_comprehensive_2023} applied a numerical fix by avoiding correction
calculations wherever the flux gradient values fall below a certain threshold. 
Tomatis \& Dall'Osso
\cite{tomatis_application_2011} developed an alternative formulation in the form of drift terms
while
